Towards Verifying the Bitcoin-S Library

### 2019-06-20 paper/paper.tex
work on intro

@@

\begin{abstract}
  We try to verify properties of the Bitcoin-S library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in subset of Scala called
  the \emph{Pure Scala Fragment}. We first try to verify the property
  that regular transactions never create new money. It turns out that
  there is too much code involved that lies outside of the supported
  fragment to make this feasible. However, in the process we uncover
  and fix a bug in Bitcoin-S. We then turn to a much simpler (and less
  interesting) property: that adding zero satoshis to a given amount
  of satoshis yields the given amount of satoshis. Here as well a
  significant part of the relevant code lies outside of the supported
  fragment. However, after a series of equivalent transformations, we
  arrive at code that we successfully verify.

@@

\subsection{Formal Verification with Stainless}
\label{sec:stainless}

kai: successor of Leon verifier \cite{DBLP:conf/ecoop/BlancKKS13,DBLP:conf/pldi/VoirolKK15,DBLP:conf/pldi/BlancK15}

Stainless is developed by "Lab for Automated Reasoning and Analysis"
(LARA) at EPFL's School of Computer and Communication Sciences.  Using
this framework we can verify the correctness of Scala programs.  The
following overview of the framework is taken from the section
\href{https://epfl-lara.github.io/stainless/intro.html}{Introduction}
of the Stainless documentation \cite{Stainless:documentation}.

Stainless statically verifies that a program satisfies a given
specification and that a program will not crash at runtime.  The
framework covers all possible inputs and finds a counter examples for
possible failures in a program which violate the given specification.

The main functions used to write a specification are \textit{require}
and \textit{ensuring}.

With \textit{require} we define a precondition of a function, which we
want to verify.  \textit{require} is placed at the beginning of the
function body.  \textit{require} specifies, if the parameter of the
function holds against a certain condition.

With \textit{ensuring}, we define the postcondition of a function.
\textit{ensuring} is placed at the end of the function after the body.
\textit{ensuring} specifies, if the return value or the result of the
function satisfies a certain condition.

On invoking Stainless, it tries to prove that the postcondition always
holds, assuming the given precondition does hold.

The following example demonstrates a simple formal specification for
the function calculating a factorial.  This is a modified example from
the section
\href{https://epfl-lara.github.io/stainless/intro.html}{Introduction}
of the Stainless documentation \cite{Stainless:documentation}, so
Stainless reports an error verifying it.
\begin{lstlisting}[language=Scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}

It recursively calculates the factorial of an integer number.
The input of the function is constrained with \textit{require} to a non-negative value.
The result should also be non-negative.
Thus, Stainless will verify whether the result of the factorial calculation is non-negative for all non-negative inputs.

Stainless can produce 3 outcomes of postcondition verification:
\textit{valid}, \textit{invalid} and \textit{unknown}.  If the
postcondition is \textit{valid} Stainless could prove that for any
inputs constrained in the precondition, the postcondition always
holds.  Reporting the postcondition as \textit{invalid} the framework
could find at least one counterexample which satisfies the
precondition but violates the postcondition.  If Stainless is unable
to prove the postcondition or find a counterexample it reports the
outcome \textit{unknown}.  In this case a timeout or an internal error
occurred.  Furthermore, Stainless checks for calls in the code to the
function with invalid arguments violating the precondition.  For
example if we invoke factorial with a negative number it reports
\textit{invalid}.

Let's return to our incorrect example with factorial.
Stainless reports the following result for the factorial verification from our example:
\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{output1.png}
	\caption{Output of Stainless verification for calculating factorial of Int number}
	\label{fig:output1}
\end{figure}

Stainless disproves the postcondition and gives the number 17 as a
counterexample.  Due to an overflow in the 32 bit Int the result
becomes a negative value.  The program would work correct by changing
the type from Int to BigInt.

Stainless supports only a subset of Scala. \todo{Anna It supports
  blabla, in particular it supports not blablabla. (check class with
  var)}

@@

Stainless verifies Scala programs using SMT (satisfiability modulo theories) solvers.

>>>>

% nicer bold
\renewcommand{\bfdefault}{b}%

% make paragraph headings bold instead of italics
\renewcommand{\paragraph}{\textbf}%

@@

\begin{abstract}
  We try to verify properties of the bitcoin-s library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in subset of Scala called
  the \emph{Pure Scala Fragment}. We first try to verify the property
  that regular transactions do not create new money. It turns out that
  there is too much code involved that lies outside of the supported
  fragment to make this feasible. However, in the process we uncover
  and fix a bug in bitcoin-s. We then turn to a much simpler (and less
  interesting) property: that adding zero satoshis to a given amount
  of satoshis yields the given amount of satoshis. Here as well a
  significant part of the relevant code lies outside of the supported
  fragment. However, after a series of equivalent transformations, we
  arrive at code that we successfully verify.

@@

\paragraph{The Stainless Verifier.} Stainless is the successor of the
Leon verifier
\cite{DBLP:conf/ecoop/BlancKKS13,DBLP:conf/pldi/VoirolKK15,DBLP:conf/pldi/BlancK15}
and is developed at EPF Lausanne
\cite{Stainless:documentation,Stainless:github}. It is intended to be
used by programmers without training in formal verification and thus
allows to write specifications in Scala and focusses on counterexample
finding in addition to proving correctness.

The following example from \cite{Stainless:documentation} demonstrates
this. Notice how a precondition is specified using the function
\emph{require} and a postcondition using \emph{ensuring}.

\begin{lstlisting}[language=Scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}

Our program happens to not satisfy the specification. An overflow in
the 32-bit Int leads to a negative result for the input 17, as
Stainless reports in Figure~\ref{fig:failed}. Changing the type from
Int to BigInt will result in a successful verification.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{output1.png}
	\caption{Output of Stainless verification for calculating factorial of Int number}
	\label{fig:failed}
\end{figure}

\paragraph{The Pure Scala Fragment.} Stainless supports only a subset
of Scala. \todo{Anna It supports blabla, in particular it supports not
  blablabla. (check class with var)}

@@
